Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-03-09 man/ecosystems/ms_ecosystems.Rmd
changes at intro

Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs of processes associated with increased intensity and duration of droughts [@Allen2010]. In fact, the frequency and severity of drought events have increased in recent decades, along with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate-change projections for the Mediterranean area forecast more frequent and severe extreme climate events [@Spinoni2017a]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005]. In a global-change context, the impact of drought on vegetation needs to be evaluated regarding the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape alteration, such as the Mediterranean region, where land-use change plays a major role in current ecosystem ecology [@Navarro2013], given that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].

The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005]. The response of rear-edge populations to climate variations could be useful in estimating the adaptation of species to predicted climatic changes [@Jump2010]. Yet it is crucial to accurately characterize the ecological rear-edge of species. This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019]. Furthermore, this characterization becomes even more critical for relict populations driven by historical land uses when they harbor high levels of intraspecific genetic diversity [@Valbuena2013].

Understanding the vegetation response to drought requires integrative and multidisciplinary strategies to assess the disturbances of forest at several scales [@Hartman2018]. The assessment of resilience to climate disturbances such as droughts provides critical information concerning the capacity of the forests to maintain their structure and render valuable ecosystem services. The drought resilience of species is better assessed when several approaches, such as dendrochronology and remote sensing, are combined. Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and annual tree-ring widths can be used to study changes in growth as a response to drought at the individiual-tree scale [@Fritts1976]. Remote sensing can be used to study the impact of drought on ecosystems and hence the stand level [*e.g.* @Zhang2013]. The combination of both approaches (remote sensing and dendroecology) has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018]. Nonetheless, it is crucial to ascertain wether the responses at the tree level differ from those at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.

In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd. (Pyrenean oak, *melojo*) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades). Our main hypothesis is that these relict populations driven by historical land use are vulnerable to climate change at their present climatic (either elevational or latitudinal) rear edge, and hence will show low resilience to extreme drought along climatic gradients at a small scale within the rear edge. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.

>>>>

Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs of more intense and longer droughts [@Allen2010]. In fact, drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate-change projections for the Mediterranean area forecast more frequent and severe extreme climate events [@Spinoni2017a]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005]. In a global-change context, the impact of drought on vegetation needs to be evaluated regarding the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape alteration, such as the Mediterranean region, where land-use change plays a major role in current ecosystem ecology [@Navarro2013], given that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].

The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution. In these cases, climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005]. The response of rear-edge populations to climate variations could be useful in estimating the adaptation of species to predicted climatic changes [@Jump2010]. Also, it is crucial to accurately characterize the ecological rear-edge of species. This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019]. Furthermore, this characterization becomes even more critical for relict populations driven by historical land uses when they harbor high levels of intraspecific genetic diversity [@Valbuena2013].

The assessment of resilience to climate disturbances such as droughts provides critical information concerning the capacity of the forests to maintain their structure and render valuable ecosystem services. The drought resilience of species is better assessed by a combination of several approaches, such as dendrochronology and remote sensing. Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and annual tree-ring widths can be used to study growth changes in response to drought among individiual trees [@Fritts1976]. Remote sensing can be used to analyze the impact of drought on ecosystems and hence on the stand [*e.g.* @Zhang2013]. The combination of both approaches (remote sensing and dendroecology) has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018]. Nonetheless, it is crucial to ascertain whether the responses at the tree level differ from those at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.

In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd. (Pyrenean oak) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades). Our main hypothesis is that these relict populations driven by historical land use are vulnerable to climate change at their present climatic (either elevational or latitudinal) rear edge, and hence will show low resilience to extreme drought along climatic gradients at a small scale within the rear edge. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.
